An Accurate Model for SARS-CoV-2 pooled RT-PCR Test Errors

### 2021-03-31 comment.tex
close to final

\begin{abstract}
Pooling is a method of simultaneously testing multiple samples for the
presence of pathogens. Pooling of SARS-CoV-2 tests is increasing in
popularity, due to its high testing throughput. A popular pooling
scheme is Dorfman pooling: test $N$ individuals simultaneously. If the
test is positive, each individual is then tested separately;
otherwise, all are declared negative. Most analyses of pooling schemes
error rates assume that including more than a single infected sample
in a pooled test does not increase the probability of a positive
outcome. We challenge this assumption with experimental data and
suggest a novel and parsimonious probabilistic model for the outcomes
of pooled tests. As an application, we analyze the false-negative rate
(i.e., the probability of a negative result for an infected
individual) of Dorfman pooling. We show that the false-negative rates
under Dorfman pooling increase when the prevalence of infection
decreases. However, low infection prevalence is exactly the condition
when Dorfman pooling achieves highest throughput efficiency. We
therefore implore the cautious use of pooling and development of
pooling strategies that consider the tests' error rates.
\end{abstract}
\newpage

@@

Studies focused on pooling for SARS-CoV-2 are in a consensus that
sample dilution effects \cite{DilutionHIV, GroupDilution} are not a
concern for pools as large as 64 individuals \cite{PoolSize30, Lion,
  DorfmanYuvalDor, DilutionCOVID}. Consequently, studies assumed that
the probability of a true-positive (the test's \emph{sensitivity})
does not depend on the number of infected samples in the pool, but
rather on the existence of at least one such sample. Thus, the
probability of a positive result in a pooled test has been (assumed)
identical for a pool with one sample from an infected individual and,
e.g., five such samples. This assumption is common in the group
testing literature \cite{Kim, OptimalDorfmanPool}, as well as in more
specific, COVID-19 focused studies \cite{Simplistic1, Simplistic2}. In
this study we challenge these commonly made assumptions and show how
using a more accurate probabilistic model affects estimation of
false-negative rates for Dorfman pooling.

@@

\subsection*{The common assumption}\label{subsec:common}
Previous studies of group testing strategies assumed that the
false-negative probability does not depend on the number of infected
samples, but merely on the existence of at least one such sample in a
pool \cite{Kim, OptimalDorfmanPool}. Current studies of pooling in the
context of SARS-CoV-2 also employ similar assumptions
\cite{Simplistic1, Simplistic2}. Specifically, these studies assume
that the probability of a negative result is the same for a pool with
a single sample coming from an infected individual and several
 samples from infected individuals. Explicitly:

@@

\subsection*{Our model}\label{subsec:ours}
Since the essence of the refuted common assumption is that
amplification of all samples occurs only once, we assume the opposite:
amplification of viral RNA succeeds or fails for each sample
independently. According to \cite{Simplistic1, Simplistic2, Kim,
  OptimalDorfmanPool}, a false-positive does not
depend on the number of negative samples in a pool. We incorporate
this assumption in our model with a small modification: We assume that
an erroneous amplification can occur in \emph{any} pool. Specifically,
it is possible that correct amplifications fails and an erroneous one
occurs simultaneously. This assumption is somewhat specific for the
current application of screening for SARS-CoV-2 via RT-PCR. For
example, cross-reactivity with other coronaviruses would have violated
this assumption. However, cross-reactivity was ruled out in
\cite{KitComparison}. These assumptions lead to our model, which is illustrated in 
Figure~\ref{fig:likelihood} and summarized in
\eqref{eq:likelihood a}.
\begin{align}
    \Pr(\data=0 | \theta) &= \spec\prod_{i=1}^N
    (1-\sens)^{\theta_i} = \spec(1-\Se)^{\sum_i\theta_i}. \label{eq:likelihood a}
\end{align}

@@

    \begin{scope}[>={Stealth[black]},
        every node/.style={fill=white},
        every edge/.style={draw=black,very thick}]
      \path [->] (A) edge node {$1-\sens$} (2,4.6);
      %\path [->] (B) edge node {$0$} (G);
      %\path [->] (D) edge node {$0$} (G);
      \path [->] (B) edge node {$1-\sens$} (2.5,4.6);
      \path [->] (F) edge node {$\spec$} (3.7,5);
      %\path [-Bar] (D) edge (4.5, 2);
    \end{scope}
  \end{tikzpicture}
  \caption{Illustration of our model. A pool contains individuals
    $\{1,2,3\}$ with state $\theta=(1,1,0)$ (i.e., individuals 1 and 2
    are infected). A negative pooled test ($\data=0$) occurs when
    three detection paths fail: A false-negative occurs for
    individuals $1$ and $2$, each with probability $(1-\sens)$.
    Additionally, no false-positive detection of SARS-CoV-2 occurs,
    with probability $\spec$. Individual $3$ is not infected and does
    not contribute to the probability of the pooled test result.}\label{fig:likelihood}
\end{figure}

\subsection*{Scheme false-negative rate}
We calculate the false-negative rate for a single \emph{infected}
individual, henceforth referred to as "Donald" under our model and
under the common assumption. We distinguish three types of
false-negative events when performing pooling. A \emph{single test}'s
false-negative is the event of a negative result upon testing Donald
separately, i.e., in an RT-PCR test without pooling. A \emph{pooled}
false-negative occurs when a pooled test containing Donald's sample
(and other samples) yields a negative result, i.e., the pooling fails
to detect at least one positive result. Lastly, a \emph{scheme}
false-negative occurs when an entire pooling scheme fails to
identify Donald as infected. Our first task is to calculate Dorfman's
scheme false-negative rate. Equivalently, what is the probability of
not identifying Donald as infected under Dorfman pooling?

Denote the prevalence of infection in the (tested) population $q$. We
assign Donald as individual $1$, so that $\theta_1=1$. Then:

@@

We calculate the probability that Donald is mistakenly identified as
not infected, henceforth referred to as the scheme's false-negative
rate and denoted $\mi$. In order to correctly identify an infected individual
as infected, both pooled and separate tests have to yield a positive
result. Thus, the scheme's false-negative rate is:
\begin{align}\label{eq:sfn}
    \begin{split}
        \mi :&= 1 - \Se\Pr(\data=1|\theta=1)\\
        %
        &= 1 - \Se\left [1 - \Sp(1-\Se)(1-q\Se))^{N-1}\right].
    \end{split}
\end{align}

@@

According to the common assumption, the scheme false-negative rate is
$1-\Se^2$. A short calculation shows that this approximation implies
the percentage increase in scheme false-negative rate is $\err{common}
:= 100\cdot \Se$.

@@

As the COVID-19 pandemic progresses, the infection prevalence
in various tested populations undergoes frequent changes. Hence, as
our results suggest, pooling strategies employed for mass testing should be
used with caution in populations where infection rates are low. Such
mass-tested populations often include air travel passengers
\cite{JTM,RobinHood}, potentially importing infections, or
presymptomatic and asymptomatic individuals, and can be crucial for
controlling outbreaks \cite{MinaScience}.

>>>>

\begin{abstract}
Pooling is a method of simultaneously testing multiple samples for the
presence of pathogens. Pooling of SARS-CoV-2 tests is increasing in
popularity, due to its high testing throughput. A popular pooling
scheme is Dorfman pooling: test $N$ individuals simultaneously. If the
test is positive, each individual is then tested separately;
otherwise, all are declared negative. Most analyses of pooling schemes
error rates assume that including more than a single infected sample
in a pooled test does not increase the probability of a positive
outcome. We challenge this assumption with experimental data and
suggest a novel and parsimonious probabilistic model for the outcomes
of pooled tests. As an application, we analyze the false-negative rate
(i.e., the probability of a negative result for an infected
individual) of Dorfman pooling. We show that the false-negative rates
under Dorfman pooling increase when the prevalence of infection
decreases. However, low infection prevalence is exactly the condition
when Dorfman pooling achieves highest throughput efficiency. We
therefore implore the cautious use of pooling and development of
pooling schemes that consider the correct tests' error rates.
\end{abstract}
\newpage

@@

Studies focused on pooling for SARS-CoV-2 are in a consensus that
sample dilution effects \cite{DilutionHIV, GroupDilution} are not a
concern, even for pools as large as 64 individuals \cite{PoolSize30,
  Lion, DorfmanYuvalDor, DilutionCOVID}. Consequently, studies assumed
that the probability of a true-positive (the test's
\emph{sensitivity}) does not depend on the number of infected samples
in the pool, but rather on the existence of at least one such
sample. Thus, the probability of a positive result in a pooled test
has been (assumed) identical for a pool with one sample from an
infected individual and, e.g., five such samples. This assumption is
common in the group testing literature \cite{Kim, OptimalDorfmanPool},
as well as in more specific, COVID-19 focused studies
\cite{Simplistic1, Simplistic2}. In this study we challenge this
commonly made assumption and show how using a more accurate
probabilistic model affects estimation of false-negative rates for
Dorfman pooling.

@@

\subsection*{The common assumption}\label{subsec:common}
Previous studies of pooling schemes assumed that the false-negative
probability does not depend on the number of infected samples, but
merely on the existence of at least one such sample in a pool
\cite{Kim, OptimalDorfmanPool}. Current studies of pooling in the
context of SARS-CoV-2 also employ a similar assumption
\cite{Simplistic1, Simplistic2}. %% Specifically, these studies assume
%% that the probability of a negative result is the same for a pool with
%% a single sample coming from an infected individual and several samples
%% from infected individuals. 
Explicitly, these studies assume:

@@

\subsection*{Our model}\label{subsec:ours}
Since the essence of the refuted common assumption is that
amplification of all samples occurs only once, we assume the opposite:
amplification of viral RNA succeeds or fails for each sample
independently. Furthermore, according to \cite{Simplistic1,
  Simplistic2, Kim, OptimalDorfmanPool}, a false-positive does not
depend on the number of negative samples in a pool either. Following
our previous reasoning, we incorporate this assumption in our model
with a small modification: We assume that a false-positive also occurs
indpendently of other amplifications. This means a false-positive can
occur in \emph{any} pool. Specifically, it is possible that every
correct amplifications fails and an erroneous one occurs
simultaneously. This assumption is somewhat specific for the current
application of screening for SARS-CoV-2 via RT-PCR. For example,
cross-reactivity with other coronaviruses would have violated this
assumption. However, cross-reactivity was ruled out in
\cite{KitComparison}. These assumptions lead to our model, which is
illustrated in Figure~\ref{fig:likelihood} and summarized in
\eqref{eq:likelihood}.
\begin{align}\label{eq:likelihood}
    \Pr(\data=0 | \theta) &= \spec\prod_{i=1}^N (1-\sens)^{\theta_i} =
    \spec(1-\Se)^{\sum_i\theta_i}.
\end{align}

@@

    \begin{scope}[>={Stealth[black]},
        every node/.style={fill=white},
        every edge/.style={draw=black,very thick}]
      \path [->] (A) edge node {$1-\sens$} (2,4.6);
      %\path [->] (B) edge node {$0$} (G);
      %\path [->] (D) edge node {$0$} (G);
      \path [->] (B) edge node {$1-\sens$} (2.5,4.6);
      \path [->] (F) edge node {$\spec$} (3.7,5);
      %\path [-Bar] (D) edge (4.5, 2);
    \end{scope}
  \end{tikzpicture}
  \caption{Illustration of our model. A pool contains individuals
    $\{1,2,3\}$ with state $\theta=(1,1,0)$ (i.e., individuals 1 and 2
    are infected). A negative pooled test ($\data=0$) occurs when
    three detection paths fail: A false-negative occurs for
    individuals $1$ and $2$, each with probability $(1-\sens)$.
    Additionally, no false-positive detection occurs, with probability
    $\spec$. Individual $3$ is not infected and does not contribute to
    the probability of the pooled test result.}\label{fig:likelihood}
\end{figure}

\subsection*{Application: scheme false-negative rate}
We calculate the false-negative rate for a single \emph{infected}
individual, henceforth referred to as "Donald", under our model and
under the common assumption. We distinguish three types of
false-negative events when performing pooling. A \emph{single test}'s
false-negative is the event of a negative result upon testing Donald
separately, i.e., in an RT-PCR test without pooling. A \emph{pooled}
false-negative occurs when a pooled test containing Donald's sample
(and other samples) yields a negative result, i.e., the pooling fails
to detect at least one positive result. Lastly, a \emph{scheme}
false-negative occurs when an entire pooling scheme fails to identify
Donald as infected. Our first task is to calculate Dorfman's scheme
false-negative rate. Equivalently, we ask: what is the probability of
not identifying Donald as infected under Dorfman pooling?

Denote the prevalence of infection in the (tested) population $q$. We
denote Donald as individual $1$, so that $\theta_1=1$. Then:

@@

We calculate the probability that Donald is mistakenly identified as
not infected, henceforth referred to as the scheme's false-negative
rate and denoted $\mi$. In order to correctly identify an infected individual
as infected, both pooled and separate tests have to yield a positive
result. Thus, the scheme's false-negative rate is:
\begin{align}\label{eq:sfn}
    \begin{split}
        \mi :&= 1 - \Se\Pr(\data=1|\theta_1=1)\\
        %
        &= 1 - \Se\left [1 - \Sp(1-\Se)(1-q\Se))^{N-1}\right].
    \end{split}
\end{align}

@@

According to the common assumption, the scheme false-negative rate is
$1-\Se^2$. A short calculation shows that this implies the percentage
increase in scheme false-negative rate is $\err{common} := 100\cdot
\Se$.

@@

As the COVID-19 pandemic progresses, the infection prevalence
in various tested populations undergoes frequent changes. Hence, as
our results suggest, pooling schemes employed for mass testing should be
used with caution in populations where infection rates are low. Such
mass-tested populations often include air travel passengers
\cite{JTM,RobinHood}, potentially importing infections, or
presymptomatic and asymptomatic individuals, and can be crucial for
controlling outbreaks \cite{MinaScience}.
